---
title: "Final Project: snRNA-seq Analysis of Human Alcohol-Dependent Brain"
author: "Milagros Galán Llario"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)
```

# 1. Load Libraries

```{r libraries}
library(Seurat)
library(harmony)
library(dplyr)
library(ggplot2)
library(patchwork)
library(SingleR)
library(scRNAseq)
library(SingleCellExperiment)
library(scuttle)
```

# 2. Load Data and Create Seurat Objects

We load the CellRanger output for each sample and create individual Seurat objects.

```{r load-data}
# Load AUD sample
aud_counts <- Read10X(
  data.dir = "/lustre/home/galanllarimm/bios668/final_project/AUD_Sample46/outs/filtered_feature_bc_matrix/"
)

aud <- CreateSeuratObject(
  counts       = aud_counts,
  project      = "AUD_Sample46",
  min.cells    = 3,
  min.features = 200
)
aud$sample    <- "AUD_Sample46"
aud$condition <- "AUD"

# Load Control sample
ctrl_counts <- Read10X(
  data.dir = "/lustre/home/galanllarimm/bios668/final_project/Control_Sample50/outs/filtered_feature_bc_matrix/"
)

ctrl <- CreateSeuratObject(
  counts       = ctrl_counts,
  project      = "Control_Sample50",
  min.cells    = 3,
  min.features = 200
)
ctrl$sample    <- "Control_Sample50"
ctrl$condition <- "Control"

cat("AUD cells before QC:    ", ncol(aud),  "\n")
cat("Control cells before QC:", ncol(ctrl), "\n")
```

# 3. Quality Control

We calculate QC metrics and visualize them to decide appropriate filtering thresholds.
In snRNA-seq from frozen postmortem brain tissue, mitochondrial content is 
naturally very low because we are sequencing nuclei, not whole cells.
A threshold of <5% MT is appropriate for this type of data.

```{r qc-metrics}
# Calculate % mitochondrial reads
aud[["percent.mt"]]  <- PercentageFeatureSet(aud,  pattern = "^MT-")
ctrl[["percent.mt"]] <- PercentageFeatureSet(ctrl, pattern = "^MT-")

# Calculate % ribosomal reads
aud[["percent.rb"]]  <- PercentageFeatureSet(aud,  pattern = "^RP[SL]")
ctrl[["percent.rb"]] <- PercentageFeatureSet(ctrl, pattern = "^RP[SL]")
```

```{r qc-plots-before, fig.height=5}
# Visualize QC metrics BEFORE filtering to decide thresholds
VlnPlot(aud,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3, pt.size = 0) +
  plot_annotation(title = "QC Metrics – AUD_Sample46 (before filtering)")

VlnPlot(ctrl,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3, pt.size = 0) +
  plot_annotation(title = "QC Metrics – Control_Sample50 (before filtering)")
```

```{r qc-scatter}
# Scatter plots to detect doublets
FeatureScatter(aud,  feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  ggtitle("AUD: nCount vs nFeature")

FeatureScatter(ctrl, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  ggtitle("Control: nCount vs nFeature")
```

```{r qc-filter}
# Filter based on visual inspection of the plots above.
# Thresholds chosen for snRNA-seq of frozen postmortem brain:
# - nFeature_RNA > 200: remove empty droplets
# - nFeature_RNA < 6000: remove likely doublets
# - percent.mt < 5: appropriate for nuclear RNA from frozen tissue

# AUD: strict threshold (good quality tissue)
aud  <- subset(aud,  subset = nFeature_RNA > 200 &
                               nFeature_RNA < 6000 &
                               percent.mt  < 5)

# Control: more permissive threshold based on visual inspection
# of violin plots — this sample shows higher MT% distribution
ctrl <- subset(ctrl, subset = nFeature_RNA > 200 &
                               nFeature_RNA < 6000 &
                               percent.mt  < 20)

cat("AUD cells after QC:    ", ncol(aud),  "\n")
cat("Control cells after QC:", ncol(ctrl), "\n")
```

```{r qc-plots-after, fig.height=5}
# Violin plots after filtering to confirm QC worked
VlnPlot(aud,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3, pt.size = 0) +
  plot_annotation(title = "QC Metrics – AUD_Sample46 (after filtering)")

VlnPlot(ctrl,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3, pt.size = 0) +
  plot_annotation(title = "QC Metrics – Control_Sample50 (after filtering)")
```

# 4. Normalization with SCTransform

SCTransform is the current standard normalization method for scRNA-seq data.
Unlike LogNormalize, SCTransform uses a regularized negative binomial regression
that corrects for technical variation (sequencing depth) while preserving
biological signal. It also automatically identifies highly variable genes (HVGs).
We regress out mitochondrial percentage to remove its confounding effect.

Note: SCTransform is run on each sample separately before merging.

```{r sctransform}
aud  <- SCTransform(aud,  vars.to.regress = "percent.mt", verbose = FALSE)
ctrl <- SCTransform(ctrl, vars.to.regress = "percent.mt", verbose = FALSE)
```

# 5. Merge Objects

We merge both samples into a single Seurat object before integration.

```{r merge}
brain <- merge(aud, y = ctrl,
               add.cell.ids = c("AUD", "CTRL"),
               project = "BrainAUD",
               merge.data = TRUE)

# Set default assay to SCT
DefaultAssay(brain) <- "SCT"

# Select variable features for integration
VariableFeatures(brain) <- SelectIntegrationFeatures(
  object.list = list(aud, ctrl),
  nfeatures = 3000
)

cat("Total cells after merge:", ncol(brain), "\n")
table(brain$sample)
```

# 6. Dimensionality Reduction (PCA)

```{r pca}
# Scale and run PCA on the merged object
brain <- ScaleData(brain, verbose = FALSE)
brain <- RunPCA(brain, npcs = 50, verbose = FALSE)
```

```{r pca-plots}
# Elbow plot to choose number of PCs to use downstream
ElbowPlot(brain, ndims = 50) +
  ggtitle("Elbow Plot – Choosing number of PCs")

# PCA plot colored by sample (before integration)
DimPlot(brain, reduction = "pca", group.by = "sample") +
  ggtitle("PCA before Harmony integration")
```

# 7. Harmony Integration

Harmony corrects for batch effects between samples while preserving biological
variation. It works on the PCA embeddings without modifying the count matrix.

```{r harmony}
brain <- RunHarmony(brain,
                    group.by.vars = "sample",
                    dims.use      = 1:30,
                    verbose       = TRUE)

# Compare before and after integration
p1 <- DimPlot(brain, reduction = "pca",     group.by = "sample") +
        ggtitle("Before Harmony (PCA)")
p2 <- DimPlot(brain, reduction = "harmony", group.by = "sample") +
        ggtitle("After Harmony")

p1 | p2
```

# 8. Clustering and UMAP

```{r cluster-umap}
# Build neighbor graph using Harmony embeddings
brain <- FindNeighbors(brain, reduction = "harmony", dims = 1:30)

# Test different resolutions to find the best number of clusters
# Human dlPFC has ~8-10 major cell types
brain <- FindClusters(brain, resolution = 0.3)
cat("Clusters at resolution 0.3:", length(unique(brain$seurat_clusters)), "\n")

brain <- FindClusters(brain, resolution = 0.5)
cat("Clusters at resolution 0.5:", length(unique(brain$seurat_clusters)), "\n")

brain <- FindClusters(brain, resolution = 0.8)
cat("Clusters at resolution 0.8:", length(unique(brain$seurat_clusters)), "\n")

# Set final resolution — adjust based on output above
brain <- FindClusters(brain, resolution = 0.5)

# Run UMAP using Harmony embeddings
brain <- RunUMAP(brain, reduction = "harmony", dims = 1:30)
```

```{r umap-plots}
# UMAP colored by cluster
p_cluster <- DimPlot(brain, reduction = "umap",
                     label = TRUE, repel = TRUE) +
               ggtitle(paste0("UMAP – Seurat clusters (",
                              length(unique(brain$seurat_clusters)),
                              " clusters)"))

# UMAP colored by sample
p_sample <- DimPlot(brain, reduction = "umap",
                    group.by = "sample") +
              ggtitle("UMAP – colored by sample")

p_cluster | p_sample
```

## 9. Find Cluster Markers

Note: `FindConservedMarkers` requires sufficient cells per cluster in each
sample. Due to sample size constraints in some clusters (e.g., Pericytes:
n=17 in AUD_Sample46), this approach did not return results. We therefore
used `FindAllMarkers`, which identifies marker genes for each cluster across
the full merged dataset. This is the standard approach for datasets with a
limited number of samples per condition.

```{r find-markers, message=FALSE, warning=FALSE}
DefaultAssay(brain) <- "SCT"
brain <- PrepSCTFindMarkers(brain)

all_markers <- FindAllMarkers(
  brain,
  only.pos        = TRUE,   # only upregulated markers per cluster
  min.pct         = 0.25,   # gene must be detected in at least 25% of cluster cells
  logfc.threshold = 0.25    # minimum log2FC threshold
)

# Top 5 markers per cluster ordered by avg_log2FC
top_markers <- all_markers %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 5) %>%
  select(cluster, gene, avg_log2FC, pct.1, pct.2, p_val_adj)

print(top_markers, n = 100)
```

```{r marker-dotplot, fig.width=14, fig.height=6}
# Top 2 markers per cluster for visualization
top2_genes <- all_markers %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 2) %>%
  pull(gene) %>%
  unique()

DotPlot(brain, features = top2_genes, dot.scale = 6) +
  RotatedAxis() +
  ggtitle("Top 2 markers per cluster (FindAllMarkers)")
```

# 10. Cell Type Annotation

## 10.1 Known brain cell type markers

```{r known-markers}
known_markers <- list(
  Excitatory_Neurons = c("SLC17A7", "CAMK2A", "NRGN"),
  Inhibitory_Neurons = c("GAD1", "GAD2", "SLC32A1"),
  Astrocytes         = c("GFAP", "AQP4", "SLC1A2"),
  Oligodendrocytes   = c("MBP", "MOG", "OLIG2"),
  OPC                = c("PDGFRA", "VCAN", "CSPG4"),
  Microglia          = c("CX3CR1", "P2RY12", "IBA1"),
  Endothelial        = c("CLDN5", "FLT1", "PECAM1"),
  Pericytes          = c("PDGFRB", "RGS5", "NOTCH3")
)

# Feature plots for each cell type
for (ct in names(known_markers)) {
  genes_present <- known_markers[[ct]][known_markers[[ct]] %in% rownames(brain)]
  if (length(genes_present) > 0) {
    p <- FeaturePlot(brain,
                     features  = genes_present,
                     reduction = "umap",
                     ncol      = length(genes_present)) &
           theme(legend.position = "right")
    print(p + plot_annotation(title = paste("Markers:", ct)))
  }
}
```

```{r dotplot-known}
all_known_present <- unlist(known_markers)
all_known_present <- all_known_present[all_known_present %in% rownames(brain)]

DotPlot(brain, features = all_known_present, dot.scale = 5) +
  RotatedAxis() +
  ggtitle("Known cell type markers across clusters")
```

### 10.2 Annotate clusters

Based on the FindAllMarkers results from Section 9, the known marker FeaturePlots,
and the SingleR validation performed in Section 10.4, we assign a cell type label
to each cluster. Four clusters were corrected relative to the initial manual annotation:

- **Cluster 0** was reassigned from Excitatory Neurons to **Oligodendrocytes**.
  This is the largest cluster in the dataset. In the human dlPFC, oligodendrocytes
  represent the most abundant glial population, making it biologically expected that
  the largest cluster corresponds to this cell type. Moreover, in a well-resolved UMAP,
  each cell type should form one spatially coherent region — the presence of "Excitatory
  Neurons" spread across multiple distal clusters (including cluster 0 and several small
  clusters on the opposite side of the UMAP) indicated a misannotation. Top markers for
  cluster 0 include canonical oligodendrocyte genes (MBP, MOG, OLIG2), consistent with
  the Oligodendrocyte label. The true Excitatory Neurons are represented by the smaller
  clusters concentrated on the left side of the UMAP (clusters 8, 10, 12, 15, 16, 19).

- **Cluster 3** was reassigned from Inhibitory Neurons to **Astrocytes**.
  Inspection of top marker genes from FindAllMarkers and the known marker FeaturePlots
  revealed expression of canonical astrocyte markers (GFAP, AQP4, SLC1A2) with no
  detectable expression of inhibitory neuron markers (GAD1, GAD2, SLC32A1). SingleR
  with the Darmanis adult brain reference also labeled this cluster as astrocytes,
  consistent with the molecular evidence. It is merged with cluster 5 into the final
  Astrocytes population.

- **Cluster 4** was reassigned from Astrocytes to OPC (top marker PDGFRA, confirmed by SingleR).

- **Cluster 19** was reassigned from Astrocytes to Excitatory Neurons (top markers
  HTR2C and NPSR1-AS1, confirmed by SingleR).

- **Cluster 18** was reassigned from Oligodendrocytes to Immune-like (top markers NKG7,
  PTPRC, RHOH, SKAP1 — consistent with a lymphoid/immune identity and labeled "hybrid"
  by SingleR with the Darmanis adult brain reference).

```{r annotate-clusters}
# Annotation vector — updated after SingleR validation in Section 10.4
cluster_annotations <- c(
  "0"  = "Oligodendrocytes",    # corrected from Excitatory Neurons per Dozmorov feedback:
                                 # largest cluster → expected to be Oligodendrocytes in dlPFC;
                                 # top markers MBP/MOG/OLIG2 confirm this;
                                 # true Excitatory Neurons are the small clusters on the left (8,10,12,15,16,19)
  "1"  = "Oligodendrocytes",
  "2"  = "Oligodendrocytes",
  "3"  = "Astrocytes",          # corrected from Inhibitory Neurons: markers show GFAP/AQP4/SLC1A2,
                                 # no GAD1/GAD2/SLC32A1; SingleR also labels as astrocytes
  "4"  = "OPC",                 # corrected from Astrocytes (PDGFRA top marker + SingleR OPC)
  "5"  = "Astrocytes",
  "6"  = "Excitatory Neurons",
  "7"  = "Inhibitory Neurons",
  "8"  = "Inhibitory Neurons",
  "9"  = "Oligodendrocytes",
  "10" = "Excitatory Neurons",
  "11" = "Endothelial",
  "12" = "Excitatory Neurons",
  "13" = "Microglia",
  "14" = "Inhibitory Neurons",
  "15" = "Excitatory Neurons",
  "16" = "Excitatory Neurons",
  "17" = "Pericytes",
  "18" = "Immune-like",         # corrected from Oligodendrocytes (NKG7/PTPRC/RHOH/SKAP1)
  "19" = "Excitatory Neurons"   # corrected from Astrocytes (HTR2C/NPSR1-AS1 — neuronal markers)
)

# Map cluster numbers to cell type labels
brain$cell_type <- unname(plyr::mapvalues(
  x            = as.character(brain$seurat_clusters),
  from         = names(cluster_annotations),
  to           = unname(cluster_annotations),
  warn_missing = FALSE
))

DimPlot(brain, reduction = "umap",
        group.by = "cell_type",
        label = TRUE, repel = TRUE) +
  ggtitle("UMAP - Annotated cell types")
```

## 10.3 Merge clusters of the same cell type

```{r merge-clusters}
Idents(brain) <- "cell_type"

DimPlot(brain, reduction = "umap",
        label = TRUE, repel = TRUE) +
  ggtitle("UMAP – Merged cell type clusters")

cat("\nCell type counts:\n")
table(brain$cell_type)

cat("\nCell type composition per sample:\n")
table(brain$cell_type, brain$sample)
```
### 10.4 Annotation validation with SingleR

To independently validate our manual cluster annotations, we used SingleR
(Aran et al., 2019) with the Darmanis adult human brain reference
(Darmanis et al., 2015), obtained from the scRNAseq Bioconductor package.
This reference contains well-characterized transcriptomic profiles for the
major CNS cell types — neurons, astrocytes, oligodendrocytes, OPCs, microglia,
and endothelial cells — making it appropriate for validating snRNA-seq data
from postmortem human brain.

The Seurat object was converted to SingleCellExperiment format and SingleR was
applied at the cluster level rather than per cell, which is more appropriate
when evaluating predefined clusters against reference labels. Because the Darmanis
reference only contains raw counts, log-normalized counts were computed prior to
classification. The resulting labels were compared with our manual annotations to
identify any discrepant clusters and inform any necessary corrections.

```{r singler-validation, message=FALSE, warning=FALSE}

# Load Darmanis adult human brain reference
darmanis <- DarmanisBrainData()

# The reference only has raw counts — add log-normalized counts for SingleR
if (!"logcounts" %in% assayNames(darmanis)) {
  logcounts(darmanis) <- log2(counts(darmanis) + 1)
}

# Convert Seurat object to SingleCellExperiment
# Note: conversion warnings about empty layers are expected with SCTransform objects
sce <- as.SingleCellExperiment(brain)

# Run SingleR at the cluster level using Darmanis as reference
pred_darmanis <- SingleR(
  test            = sce,
  ref             = darmanis,
  labels          = darmanis$cell.type,
  clusters        = sce$seurat_clusters,
  assay.type.test = "logcounts",
  assay.type.ref  = "logcounts"
)

# Build comparison table: manual annotation vs SingleR prediction per cluster
manual_df <- data.frame(
  cluster = levels(brain$seurat_clusters),
  manual  = sapply(levels(brain$seurat_clusters), function(cl) {
    names(sort(table(brain$cell_type[brain$seurat_clusters == cl]),
               decreasing = TRUE))[1]
  })
)

darmanis_df <- data.frame(
  cluster         = rownames(pred_darmanis),
  singler_label   = pred_darmanis$labels,
  singler_pruned  = pred_darmanis$pruned.labels
)

comparison <- merge(manual_df, darmanis_df, by = "cluster")
comparison <- comparison[order(as.numeric(as.character(comparison$cluster))), ]

cat("\nCluster annotation comparison: manual vs SingleR (Darmanis adult brain)\n")
print(comparison, row.names = FALSE)

# Save table
write.csv(comparison,
          file = "/lustre/home/galanllarimm/bios668/final_project/singler_comparison.csv",
          row.names = FALSE)
cat("\nComparison table saved.\n")
```

```{r singler-umap, fig.width=12, fig.height=5}
# Add Darmanis SingleR predictions to Seurat object for visual comparison
brain$singler_label <- pred_darmanis$labels[
  match(as.character(brain$seurat_clusters), rownames(pred_darmanis))
]

# Side-by-side UMAP: manual annotation vs SingleR
p_manual <- DimPlot(brain, reduction = "umap",
                    group.by = "cell_type",
                    label = TRUE, repel = TRUE, pt.size = 0.3) +
  ggtitle("Manual annotation") +
  theme(legend.position = "bottom")

p_singler <- DimPlot(brain, reduction = "umap",
                     group.by = "singler_label",
                     label = TRUE, repel = TRUE, pt.size = 0.3) +
  ggtitle("SingleR annotation (Darmanis adult brain)") +
  theme(legend.position = "bottom")

p_manual | p_singler
```

# 11. Final Annotated UMAP Split by Sample

```{r final-umap, fig.width=14, fig.height=7}
DimPlot(brain,
        reduction = "umap",
        group.by  = "cell_type",
        split.by  = "sample",
        label     = TRUE,
        repel     = TRUE,
        pt.size   = 0.3) +
  ggtitle("Final Annotated UMAP – AUD vs Control") +
  theme(legend.position = "bottom")
```

```{r final-barplot, fig.height=5}
prop_df <- as.data.frame(table(brain$cell_type, brain$sample))
colnames(prop_df) <- c("CellType", "Sample", "Count")

ggplot(prop_df, aes(x = Sample, y = Count, fill = CellType)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Cell type proportions per sample",
       y = "Proportion", x = "Sample") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 12. Save Seurat Object

```{r save}
saveRDS(brain,
        file = "/lustre/home/galanllarimm/bios668/final_project/brain_final.rds")
cat("Object saved successfully!\n")
```

# 13. Session Info

```{r session}
sessionInfo()
```
