---
title: "Final Project – Step 6: Differential Expression & Functional Enrichment"
author: "Milagros Galán Llario"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)
```

# 1. Load Libraries

```{r libraries}
library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)

# For functional enrichment
library(clusterProfiler)
library(org.Hs.eg.db)   # Human gene annotation database
library(enrichplot)
library(ggrepel)        # For volcano plot labels
```

# 2. Load Seurat Object

We load the previously processed and annotated Seurat object from Step 5.
This object already contains QC-filtered, SCTransform-normalized, Harmony-integrated,
and cell-type-annotated data for the two samples (AUD_Sample46 and Control_Sample50).

```{r load-object}
brain <- readRDS(
  "/lustre/home/galanllarimm/bios668/final_project/brain_final.rds"
)

cat("Total cells loaded:", ncol(brain), "\n")
cat("Samples:\n")
print(table(brain$sample))
cat("\nCell type distribution:\n")
print(sort(table(brain$cell_type), decreasing = TRUE))
```

# 3. Identify Top 3 Largest Cell Type Clusters

We identify the three most abundant cell types across both samples.
Differential expression will be performed within each of these cell types,
comparing AUD versus Control nuclei.

```{r top3-celltypes}
# Count cells per cell type and sort
ct_counts <- sort(table(brain$cell_type), decreasing = TRUE)

cat("Cell type sizes (all):\n")
print(ct_counts)

# Select top 3
top3_celltypes <- names(ct_counts)[1:3]
cat("\nTop 3 largest cell types selected for DE analysis:\n")
for (i in seq_along(top3_celltypes)) {
  cat(sprintf("  %d. %s (n = %d)\n", i, top3_celltypes[i], ct_counts[top3_celltypes[i]]))
}
```

# 4. Differential Expression Analysis

## 4.1 Setup

For differential expression between conditions (AUD vs Control), we need
to use the **RNA assay** rather than SCT, as recommended by Seurat for
cross-sample DE testing. We also log-normalize the RNA assay to ensure
proper scale for the Wilcoxon test.

**Statistical note**: With only one sample per condition (n=1 AUD, n=1 Control),
pseudobulk approaches (which require biological replicates) are not applicable here.
We therefore use the Wilcoxon rank-sum test (Seurat default) at the single-cell level.
Results should be interpreted with caution and treated as exploratory, as individual
cells within one sample are not independent biological replicates.

```{r de-setup}
# Switch to RNA assay for DE analysis
DefaultAssay(brain) <- "RNA"

# Log-normalize the RNA assay (required for Wilcoxon test)
brain <- NormalizeData(brain, verbose = FALSE)

# Create a combined label: cell_type + condition (e.g. "Oligodendrocytes_AUD")
brain$celltype.condition <- paste(brain$cell_type, brain$condition, sep = "_")

cat("Combined identity labels (first 10):\n")
print(head(sort(unique(brain$celltype.condition)), 10))
```

## 4.2 Run FindMarkers for Each Top Cell Type

For each of the top 3 cell types, we compare AUD (ident.1) vs Control (ident.2).
Positive log2FC values indicate upregulation in AUD; negative values indicate
upregulation in Control.

**Mitochondrial gene exclusion**: mitochondrial genes (those with names starting
with "MT-") are excluded from the DE analysis prior to testing. In snRNA-seq data
from frozen postmortem tissue, mitochondrial gene expression is heavily influenced
by the quality of nuclear isolation and the degree of cytoplasmic contamination,
which can differ substantially between samples due to technical rather than biological
reasons. In this dataset, Control_Sample50 showed a broader mitochondrial % distribution
during QC, which required a less stringent threshold than AUD_Sample46. As a consequence,
mitochondrial genes dominated the DE results and obscured biologically meaningful
differences. Excluding them upstream of FindMarkers ensures that the DE results
reflect genuine transcriptional differences between AUD and Control rather than
technical artifacts of sample quality.

```{r run-de, message=FALSE, warning=FALSE}
# Set identity to the combined cell type + condition label
Idents(brain) <- "celltype.condition"

# Define non-mitochondrial genes to test
# MT- genes are excluded because their differential expression in this dataset
# reflects QC differences between samples (Control_Sample50 had broader MT%
# distribution), not true biological differences between AUD and Control.
non_mito_genes <- rownames(brain)[!grepl("^MT-", rownames(brain))]
cat(sprintf("Total genes in object: %d | After MT- exclusion: %d | MT- genes removed: %d\n",
            nrow(brain), length(non_mito_genes), nrow(brain) - length(non_mito_genes)))

# Run DE for each of the top 3 cell types
de_results <- list()

for (ct in top3_celltypes) {
  
  ident1 <- paste0(ct, "_AUD")
  ident2 <- paste0(ct, "_Control")
  
  # Check that both groups exist and have enough cells
  n1 <- sum(brain$celltype.condition == ident1)
  n2 <- sum(brain$celltype.condition == ident2)
  
  cat(sprintf("\n--- %s: AUD n=%d | Control n=%d ---\n", ct, n1, n2))
  
  if (n1 < 10 || n2 < 10) {
    cat("  Skipping: insufficient cells in one group (<10).\n")
    next
  }
  
  de <- FindMarkers(
    brain,
    ident.1          = ident1,
    ident.2          = ident2,
    test.use         = "wilcox",   # Wilcoxon rank-sum test (appropriate for 1 sample/condition)
    features         = non_mito_genes, # Exclude MT- genes (technical artifact in this dataset)
    min.pct          = 0.10,       # Gene must be detected in >=10% of cells in either group
    logfc.threshold  = 0.1,        # Lower threshold to capture modest effects
    assay            = "RNA",
    verbose          = FALSE
  )
  
  # Add gene name and cell type as columns
  de$gene      <- rownames(de)
  de$cell_type <- ct
  
  # Add significance classification
  de$significance <- "Not significant"
  de$significance[de$p_val_adj < 0.05 & de$avg_log2FC >  0.25] <- "Up in AUD"
  de$significance[de$p_val_adj < 0.05 & de$avg_log2FC < -0.25] <- "Up in Control"
  
  de_results[[ct]] <- de
  
  n_sig <- sum(de$significance != "Not significant")
  n_up  <- sum(de$significance == "Up in AUD")
  n_dn  <- sum(de$significance == "Up in Control")
  cat(sprintf("  Total tested: %d | Significant: %d (Up AUD: %d | Up Control: %d)\n",
              nrow(de), n_sig, n_up, n_dn))
}
```

## 4.3 Summary Table of Significant DE Genes

```{r de-summary}
# Combine all results and show significant genes
all_de <- bind_rows(de_results)

sig_de <- all_de %>%
  filter(significance != "Not significant") %>%
  arrange(cell_type, p_val_adj) %>%
  dplyr::select(cell_type, gene, avg_log2FC, pct.1, pct.2, p_val_adj, significance)

cat("Significant DE genes (p_adj < 0.05, |log2FC| > 0.25):\n")
print(head(sig_de, n = 50))

# Save full DE results to CSV
write.csv(all_de,
          file = "/lustre/home/galanllarimm/bios668/final_project/DE_results_top3_celltypes.csv",
          row.names = FALSE)
cat("\nFull DE results saved.\n")
```

# 5. Volcano Plots

Volcano plots show the relationship between effect size (log2FC, x-axis) and
statistical significance (-log10 adjusted p-value, y-axis). Each point is one gene.
Red = upregulated in AUD, blue = upregulated in Control, grey = not significant.

```{r volcano-plots, fig.height=6, fig.width=8}
for (ct in names(de_results)) {
  
  de <- de_results[[ct]]
  
  # Label top 15 most significant genes
  top_genes <- de %>%
    filter(significance != "Not significant") %>%
    arrange(p_val_adj) %>%
    slice_head(n = 15)
  
  p <- ggplot(de, aes(x = avg_log2FC,
                      y = -log10(p_val_adj + 1e-300),
                      color = significance)) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_manual(
      values = c("Up in AUD"      = "#E64B35",
                 "Up in Control"  = "#4DBBD5",
                 "Not significant" = "grey70")
    ) +
    geom_text_repel(
      data        = top_genes,
      aes(label   = gene),
      size        = 3,
      max.overlaps = 20,
      color       = "black"
    ) +
    geom_vline(xintercept = c(-0.25, 0.25), linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = -log10(0.05),   linetype = "dashed", color = "grey40") +
    labs(
      title    = paste0("Volcano Plot – ", ct),
      subtitle = "AUD vs Control | Wilcoxon test | padj < 0.05 & |log2FC| > 0.25",
      x        = "Average log2 Fold Change (AUD / Control)",
      y        = "-log10(adjusted p-value)",
      color    = "Direction"
    ) +
    theme_classic() +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 10, color = "grey40"),
      legend.position = "top"
    )
  
  print(p)
}
```

# 6. Functional Enrichment Analysis

## 6.1 Overview

For each of the top 3 cell types, we run two types of enrichment analysis on
the significant DE genes:

- **GO Biological Process (ORA)**: identifies which biological processes (Gene Ontology
  terms) are over-represented in our gene list compared to all expressed genes.
- **KEGG Pathway (ORA)**: identifies which signaling/metabolic pathways are
  over-represented.

Both use **Over-Representation Analysis (ORA)**, which applies a hypergeometric
test (equivalent to Fisher's exact test) asking: *"Are genes from pathway X
more common in my DE list than expected by chance?"*

We perform enrichment separately for **upregulated** (AUD > Control) and
**downregulated** (AUD < Control) gene sets, since mixing both directions
can obscure biological interpretation.

## 6.2 GO Biological Process Enrichment

```{r go-enrichment, fig.height=8, fig.width=10}
go_results <- list()

for (ct in names(de_results)) {
  
  de <- de_results[[ct]]
  
  # Universe = all genes tested in this cell type
  universe_genes <- de$gene
  
  # Up in AUD
  up_genes <- de %>%
    filter(significance == "Up in AUD") %>%
    pull(gene)
  
  # Down in AUD (= up in Control)
  dn_genes <- de %>%
    filter(significance == "Up in Control") %>%
    pull(gene)
  
  go_results[[ct]] <- list()
  
  for (direction in c("Up_AUD", "Up_Control")) {
    
    gene_set <- if (direction == "Up_AUD") up_genes else dn_genes
    
    if (length(gene_set) < 5) {
      cat(sprintf("\n%s – %s: fewer than 5 genes, skipping GO enrichment.\n", ct, direction))
      next
    }
    
    ego <- tryCatch({
      enrichGO(
        gene          = gene_set,
        universe      = universe_genes,
        OrgDb         = org.Hs.eg.db,
        keyType       = "SYMBOL",
        ont           = "BP",             # Biological Process
        pAdjustMethod = "BH",
        pvalueCutoff  = 0.05,
        qvalueCutoff  = 0.2,
        readable      = TRUE
      )
    }, error = function(e) {
      cat(sprintf("  GO enrichment failed for %s %s: %s\n", ct, direction, e$message))
      NULL
    })
    
    if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
      go_results[[ct]][[direction]] <- ego
      
      cat(sprintf("\n%s – %s: %d enriched GO:BP terms\n",
                  ct, direction, nrow(as.data.frame(ego))))
      
      # Dotplot
      p <- dotplot(ego, showCategory = 15, font.size = 9) +
        ggtitle(paste0("GO:BP – ", ct, " (", gsub("_", " ", direction), ")")) +
        theme(plot.title = element_text(face = "bold"))
      print(p)
      
    } else {
      cat(sprintf("\n%s – %s: no enriched GO:BP terms found.\n", ct, direction))
    }
  }
}
```

## 6.3 KEGG Pathway Enrichment

KEGG requires Entrez gene IDs instead of gene symbols. We use `bitr()` from
clusterProfiler to convert gene symbols to Entrez IDs before running enrichKEGG().

```{r kegg-enrichment, fig.height=7, fig.width=10}
kegg_results <- list()

for (ct in names(de_results)) {
  
  de <- de_results[[ct]]
  kegg_results[[ct]] <- list()
  
  # Convert universe to Entrez IDs
  universe_conv <- tryCatch(
    bitr(de$gene, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db),
    error = function(e) NULL
  )
  
  if (is.null(universe_conv)) next
  universe_entrez <- universe_conv$ENTREZID
  
  for (direction in c("Up_AUD", "Up_Control")) {
    
    gene_set <- de %>%
      filter(significance == if (direction == "Up_AUD") "Up in AUD" else "Up in Control") %>%
      pull(gene)
    
    if (length(gene_set) < 5) {
      cat(sprintf("\n%s – %s: fewer than 5 genes, skipping KEGG enrichment.\n", ct, direction))
      next
    }
    
    # Convert gene symbols to Entrez IDs
    gene_conv <- tryCatch(
      bitr(gene_set, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db),
      error = function(e) NULL
    )
    
    if (is.null(gene_conv) || nrow(gene_conv) == 0) next
    gene_entrez <- gene_conv$ENTREZID
    
    kk <- tryCatch({
      enrichKEGG(
        gene          = gene_entrez,
        universe      = universe_entrez,
        organism      = "hsa",            # Homo sapiens
        keyType       = "ncbi-geneid",
        pAdjustMethod = "BH",
        pvalueCutoff  = 0.05,
        qvalueCutoff  = 0.2
      )
    }, error = function(e) {
      cat(sprintf("  KEGG enrichment failed for %s %s: %s\n", ct, direction, e$message))
      NULL
    })
    
    if (!is.null(kk) && nrow(as.data.frame(kk)) > 0) {
      
      # Make gene names readable (convert Entrez → Symbol in output)
      kk <- setReadable(kk, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
      kegg_results[[ct]][[direction]] <- kk
      
      cat(sprintf("\n%s – %s: %d enriched KEGG pathways\n",
                  ct, direction, nrow(as.data.frame(kk))))
      
      # Dotplot
      p <- dotplot(kk, showCategory = 15, font.size = 9) +
        ggtitle(paste0("KEGG – ", ct, " (", gsub("_", " ", direction), ")")) +
        theme(plot.title = element_text(face = "bold"))
      print(p)
      
    } else {
      cat(sprintf("\n%s – %s: no enriched KEGG pathways found.\n", ct, direction))
    }
  }
}
```

## 6.4 Gene-Concept Network (cnetplot)

The cnetplot shows which specific genes drive each enriched term,
and which genes belong to multiple terms simultaneously.

```{r cnetplot, fig.height=9, fig.width=11}
for (ct in names(go_results)) {
  for (direction in names(go_results[[ct]])) {
    
    ego <- go_results[[ct]][[direction]]
    
    if (is.null(ego) || nrow(as.data.frame(ego)) == 0) next
    
    # Get log2FC values for color coding
    de <- de_results[[ct]]
    lfc_vec <- setNames(de$avg_log2FC, de$gene)
    
    p <- cnetplot(
      ego,
      showCategory  = 5,
      foldChange    = lfc_vec,
    ) +
      ggtitle(paste0("Gene-Concept Network – ", ct,
                     " (", gsub("_", " ", direction), ")")) +
      theme(plot.title = element_text(face = "bold"))
    
    print(p)
  }
}
```

# 7. Save Enrichment Results

```{r save-enrichment}
# Save GO results
for (ct in names(go_results)) {
  for (direction in names(go_results[[ct]])) {
    ego <- go_results[[ct]][[direction]]
    if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
      fname <- paste0(
        "/lustre/home/galanllarimm/bios668/final_project/GO_BP_",
        gsub(" ", "_", ct), "_", direction, ".csv"
      )
      write.csv(as.data.frame(ego), file = fname, row.names = FALSE)
      cat("Saved:", fname, "\n")
    }
  }
}

# Save KEGG results
for (ct in names(kegg_results)) {
  for (direction in names(kegg_results[[ct]])) {
    kk <- kegg_results[[ct]][[direction]]
    if (!is.null(kk) && nrow(as.data.frame(kk)) > 0) {
      fname <- paste0(
        "/lustre/home/galanllarimm/bios668/final_project/KEGG_",
        gsub(" ", "_", ct), "_", direction, ".csv"
      )
      write.csv(as.data.frame(kk), file = fname, row.names = FALSE)
      cat("Saved:", fname, "\n")
    }
  }
}

cat("\nAll enrichment results saved successfully.\n")
```

# 8. Session Info

```{r session}
sessionInfo()
```
